

Olympic Data Analysis (Since 2000)
========================================================

**Jason Ng, 
Jesus Martin,
Jakob Rieckh,
Nihar R Prasad,
Priya Sankarlingam,
**


The Data
--------------------------------------------

We took the sample data for Olympics (both summer and winter) from Tableau website. This sample data set has medals (all types) and athletes from each Olympic Games since the 2000 Games in Sydney, Australia. So, the number of data set for each type of Olympics (summer and winter) is four.

We sanity checked the data for several factors including data being in metric form and our empirical contextual knowledge of Olympics.


Objective
--------------------------------------------

We set out to analyse the given data in terms of: 

• country ranking per sport

•	average age of winning a medal

Our goal is to have a better understanding on which countries are better at which sports and which countries are better at Olympics in general.

Then, we want to predict the number of medals that each country will win in the next Summer Olympics, based on how many medals won in the last 4 Summer Olympic games. In order to do that, we will use a linear regression. 

We are also planning to develop an application to come-up with more flexibility over the data and more customized conclusions.


Data Validation
--------------------------------------------

We extracted the data from the Tableau website. At the beginningg, we thought it would be a good cource of information to do data analysis but the main challenge that we have faced is format the data in a proper way to get it suitable for our analysis.

We have done trend analysis to estimate the model. We have analysed trend for top country for each sport to further predict a data point for the next Olympics. This has been divided by the two types of Olympics. We further assessed performance on first validation data to see if 2012 figures match the predictive model that we have developed.
We reiterated the predictive model and finally were able to find a predictor which was closes to London 2012 figures.


Final Results
--------------------------------------------

•  Our analysis showed that since 2000,  United States has been the best country in getting Olympic medals combined in all sports.
•	Our prediction is that in the next Olympics,  would be getting the highest number of total medals.
•	Further, by each sport, we have the following table for top countries per sport.


```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}
table1 <- summarise(group_by(melt(ProjectData, id=c(3,6,10), measure=c(10)),Country, Sport), sum(value))
names(table1)[3] <- "Medals"
max <- aggregate(table1$Medals, list(Sport= table1$Sport), max)
names(max)[2] <- "Medals"
table1max<- merge(max,table1)
table1print<- table1max[,c(1,3)]
names(table1print)[2] <- "Top Country"
m1<-gvisTable(table1print,options=list(showRowNumber=TRUE,width=900, height=min(300,27*50),allowHTML=TRUE,page='disable'))
print(m1,'chart')
````

Based on these results, we can infer the following:


For mountain sports like alpine skiing, countries like Austria and Switzerland are on the top.

For athletics, there was a diverse presence from three continents - USA, Russia and Jamaica
Cuba has a varied presence in strength sports like Boxing and Baseball

For table tennis and badminton, traditional players like China and South Korea have continued to be strong

Russia continues to be one of the strongest countries in Weightlifting and Wrestling

China is also strong in Badminton or Table Tennis but also in Shooting and many other sports

USA and Australia continue to dominate swimming.


Based on the following ranking of countries by the number of countries in which are leaders in medals 
```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}
countries <- data.frame(table(table1print[,2]))
scountries<- countries[order(-countries$Freq),]
tcountries<- scountries[which(scountries$Freq > 0),]
names(tcountries)[2]<- "number of sports"
names(tcountries)[1]<- "Country"
m2<-gvisTable(tcountries,options=list(showRowNumber=FALSE,width=500, height=min(200,27*20),allowHTML=TRUE,page='disable'))
print(m2,'chart')
```


Overall, USA continues to lead by far with China, Germany and Great Britain following and Russia a close 4th, which we can visualize in the following map:
```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}

geo <- gvisGeoChart(tcountries,locationvar="Country", 
                 colorvar="number of sports",
                 options=list(projection="kavrayskiy-vii",width=700, height=500, page='disable'))
print(geo,'chart')
```


```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}
ProjectData.m <- melt(ProjectData, id=c(6), measure=c(2))
table2 <- cast(ProjectData.m, Sport ~ variable, mean)
table2 <- table2[which(table2$Age > 0),]
mean(ProjectData$Age, na.rm=TRUE)
mintable2 <- min(table2$Age)
```

Average age by sport
--------------------------------------------
We took the age of the medal winners in different sports and averaged it to have a metric that would help us at what age do athletes get medals depending on the sport. 
```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}
Bar <- gvisBarChart(table2,
                    options=list(legend='none', width=900, height=1600))
print(Bar,'chart')
````
Based on age based analysis, we came up with following major insights:


The average age of an Olympian winning a medal is `r mean(ProjectData$Age, na.rm=TRUE)` years

The youngest people to win medals are in Gymnasts - particularly, Rythmic Gymnasts (`r min(table2$Age)` years)

Also, with water sports the average age is quite low at around 22.5 years

However, as we thought, with practice sports like Shooting, Curling and Equestrian the average age is high 
above 30 years with Equestrian having the highest average age at 37.9 years

Also, by law of averages, team sports like Hockey, Basketball and Volleyball have average age near the total average of 26.4 years.



Number of medals for next summer Olympics?
--------------------------------------------

Task Description
For one specific country (in our example we picked the United States), we wanted to use the total number of medals won at past Olympic games to predict the metal yield in 2016. We applied a linear regression model to conduct this forecast. Since the number of medals could vary significantly between winter and summer games, we selected only medals won at summer Olympics.
 
Approach
1)      Convert data into a Pivot Table

2)      Filter out the medals won by the US at summer Olympics

3)      Run the regression (independent variables: year dates -> dependent variables: sum of medals)

4)      Predict the number of medals for 2016 based on the calculated regression coefficients



```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}

# Step 4: Estimate the number of medials for 2016
medalForcast = coeffs[1] + coeffs[2]*2016 
if (medalForcast<0) medalForcast <- 0
if (coeffs[2]>0) trend <- "positive"
if (coeffs[2]<0) trend <- "negative"
newYears <- c(years, 2016)
newMedals <- c(selectedSummer, medalForcast)
plot(newYears, newMedals, xaxt = "n", xlab="Summer Olympics",ylab="Number of Medals over all Athletes")
axis(side=1, at=newYears)
points(2016, medalForcast, col="red")
title("Scatter Plot for Medals Regression")
abline(fit)
````


The number of medals expected for `r reg_country` in 2016 is `r medalForcast`

**Conclusion**

The number of medals won by the `r reg_country` team at the last four summer Olympics vary between `r min(selectedSummer)` and `r max(selectedSummer)`. The overall trend is `r trend`, resulting in a regression function with a slope of `r coeffs[2]`. Extrapolating this trend to 2016, gives us a prediction of `r medalForcast`  medals for `r reg_country` in Rio de Janeiro.
 
**Limitations**

1)      The data source lists all athletes that won Olympic medals – regards if it’s an individual or team sport. This means that team sports have a stronger influence on the regression outcome than sports with only one athlete.

2)      Due to the underlying mechanics of a linear regression all Olympic games influence the forecast for 2016 equally. For a more accurate prediction it would be reasonable to put more weight on the recent years (2012, 2008) at the expense of the earlier years (2004, 2000).

3)      For the sake of this exercise we assumed that the number of sports and team sizes stay constant over the various Olympic games.




